---
title: "acropora_functional_enrichment"
author: "LM Gorman, AS Huffmyer"
date: "2025-03-28"
output:
  html_document:
    toc: yes
    toc_depth: 6
    toc_float: yes
  pdf_document:
    keep_tex: yes
editor_options: 
  chunk_output_type: console
---




# Set up 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# The following setting is important, do not omit.
options(stringsAsFactors = FALSE) #Set Strings to character
```

```{r}
if ("tidyverse" %in% rownames(installed.packages()) == 'FALSE') install.packages('tidyverse') 
if ("genefilter" %in% rownames(installed.packages()) == 'FALSE') install.packages('genefilter') 
if ("DESeq2" %in% rownames(installed.packages()) == 'FALSE') install.packages('DESeq2') 
if ("RColorBrewer" %in% rownames(installed.packages()) == 'FALSE') install.packages('RColorBrewer') 
#if ("WGCNA" %in% rownames(installed.packages()) == 'FALSE') install.packages('WGCNA') 
if ("flashClust" %in% rownames(installed.packages()) == 'FALSE') install.packages('flashClust') 
if ("gridExtra" %in% rownames(installed.packages()) == 'FALSE') install.packages('gridExtra') 
if ("ComplexHeatmap" %in% rownames(installed.packages()) == 'FALSE') install.packages('ComplexHeatmap') 
if ("goseq" %in% rownames(installed.packages()) == 'FALSE') install.packages('goseq') 
if ("dplyr" %in% rownames(installed.packages()) == 'FALSE') install.packages('dplyr') 
if ("clusterProfiler" %in% rownames(installed.packages()) == 'FALSE') install.packages('clusterProfiler') 
if ("pheatmap" %in% rownames(installed.packages()) == 'FALSE') install.packages('pheatmap') 
if ("magrittr" %in% rownames(installed.packages()) == 'FALSE') install.packages('magrittr') 
if ("rtracklayer" %in% rownames(installed.packages()) == 'FALSE') BiocManager::install("rtracklayer")
if ("GenomicRanges" %in% rownames(installed.packages()) == 'FALSE') BiocManager::install("GenomicRanges")
if ("plyranges" %in% rownames(installed.packages()) == 'FALSE') BiocManager::install("plyranges")
if ("GSEABase" %in% rownames(installed.packages()) == 'FALSE') BiocManager::install("GSEABase")
#if ("GOSim" %in% rownames(installed.packages()) == 'FALSE') BiocManager::install("GOSim")
if ("stats" %in% rownames(installed.packages()) == 'FALSE') install.packages("stats")
if ("ggdendro" %in% rownames(installed.packages()) == 'FALSE') BiocManager::install("ggdendro")
if ("GO.db" %in% rownames(installed.packages()) == 'FALSE') BiocManager::install("GO.db")
if ("rrvgo" %in% rownames(installed.packages()) == 'FALSE') BiocManager::install("rrvgo")
if ("rtracklayer" %in% rownames(installed.packages()) == 'FALSE') BiocManager::install("rtracklayer")

#BiocManager::install("org.Ce.eg.db", force=TRUE) #install if needed 
library("rtracklayer")
library("tidyverse")
library("genefilter")
library("DESeq2")
library("RColorBrewer")
#library("WGCNA")
library("flashClust")
library("gridExtra")
library("ComplexHeatmap")
library("goseq")
library("dplyr")
library("clusterProfiler")
library("pheatmap")
library("magrittr")
library("rtracklayer")
library("GenomicRanges")
library("plyranges")
library("GSEABase")
#library("GOSim")
library("stats")
library("ggdendro")
library("GO.db")
library("rrvgo")
library("cowplot")
```

# Read in data files 

1. Read in file with vst transformed counts of all genes detected and kept after filtering. 
```{r Read in all genes file}
#Need to work out which file this is from acropora_DEG script
all_genes_acr<-read.csv("data/all_genes_acr.csv", col.names = c("gene_id","trim.321ra.gtf","trim.331ra.gtf","trim.336r.gtf","trim.370r.gtf","trim.380r.gtf","trim.410r.gtf","trim.414r.gtf","trim.419r.gtf","trim.468r.gtf","trim.512r.gtf","trim.549r.gtf","trim.568r.gtf","trim.571r.gtf","trim.581r.gtf","trim.586r.gtf"))
```

2. Read in file of DEG's between eaten and control
```{r read in file of DEGs}
acr_eatenvscontrol_DEG_list<-read.csv("data/DEG_acr.csv")
acr_eatenvscontrol_DEG_list$contrast<-factor(acr_eatenvscontrol_DEG_list$contrast, levels=c("eaten", "control"))
```

# Read in annotation files 
Will use the genome annotations done in the google drive for *A. hyacinthus* (López-Nandam et al., 2023; https://royalsocietypublishing.org/doi/10.1098/rspb.2022.1766). This paper used MAKER2 for genome annotation. I think the annotation file is called: Ahyacinthus_transcripts_b2g-annotated_blast-nr-results.txt (check with Jill/Ariana this is correct)

NB:Many scientists use funannotate package in R (https://funannotate.readthedocs.io/en/latest/index.html) to run functional enrichment of genomes.
funannotate - Script functionally annotates the results from funannotate predict.  It pulls annotation from PFAM, InterPro, EggNog, UniProtKB, MEROPS, CAZyme, and GO ontology.


Conduct functional enrichment using goseq and rrvgo. Compare against all genes detected in the dataset that we have annotation information for.

I need to use the genome annotations from google drive to compare against my geneset. Below is Ariana's script from https://github.com/AHuffmyer/EarlyLifeHistory_Energetics/blob/master/Mcap2020/Scripts/TagSeq/DESeq2_Mcap_V3.Rmd. I need to paste my relevant documents and data. Ariana clustered her data during this - do I need to do that too?

```{r annotation files}
Ahya_annot <- read.delim("data/Ahyacinthus_transcripts_b2g-annotated_blast-nr-results.txt",  header=TRUE, na.strings = "---NA---")
dim(Ahya_annot)
#Need to remove RA at the end of all the SeqNames so it matches to gene in acr_eatenvscontrol_DEG
Ahya_annot_seq_name <- Ahya_annot %>%
  mutate(SeqName = str_remove(SeqName, "-RA")) 
```
This functional annotation has 27110 genes. 
SeqNames now match the gene vairable in both files.
```{r}
# look at variable names
names(Ahya_annot)
#remove sample column name 
probes = all_genes_acr$gene_id
seq_name = Ahya_annot_seq_name$SeqName

probes_in_annot = probes %in% seq_name

seqname_in_probes_df <- as.data.frame(seq_name %in% probes, row.names = Ahya_annot_seq_name$SeqName)

View(probes_in_annot) #match genes (SeqName)to genes in annotation file.

# The following is the number of probes without annotation 
sum(probes_in_annot == FALSE) #0
sum(probes_in_annot == TRUE) #18258 = all genes found in annotation file - yayy!
seqname_in_probes = seq_name %in% probes
sum(seqname_in_probes == FALSE) #8852
sum(seqname_in_probes == TRUE)#18258
#view the genes that do not have a match in the annotation file
seqname_in_probes_df <- as.data.frame(seq_name %in% probes, row.names = Ahya_annot_seq_name$SeqName) 
seqname_in_probes_df$RowNames <- rownames(seqname_in_probes_df)

#Create new variable saying when genes are missing from the dataframe
seqname_in_probes_df <- seqname_in_probes_df %>%
  mutate(missing_genes =
         case_when(
           seqname_in_probes_df == "FALSE" ~ "YES"
           )
         )
print(seqname_in_probes_df)

#Now use this column and row names to filter the larger data frame of gene annotations (Ahya_annot_seq_name) to only include our genes found in our A. hyacinthus RNASeq
missing_gene_list <- seqname_in_probes_df %>%
  filter(missing_genes == "YES")
dim(missing_gene_list) #8852 as expected - great
#Now create a filtered gene set taking away the missing_gene_list from the annotated gene list
Ahya_annot_seq_name_filtered <- Ahya_annot_seq_name %>% 
  filter(!(SeqName %in% missing_gene_list$RowNames))
dim(Ahya_annot_seq_name_filtered)
```

There are 8852 of our genes that were expressed that are not present in the annotation file.  

Reduced annotation file to only contain genes detected in our dataset.  

Now file only contains the 18258 genes after filtering. This is 18258 out of the 27110 genes in our dataset. 

Add in length to the annotation file.   

Get gene length information.  
```{r}
#import file
#Not sure which file to import into R? there are .gff and .gff3 files both in the /data directory.
gff <- rtracklayer::import("data/Ahyacinthus.coding.gff3") #if this doesn't work, restart R and try again 

transcripts <- subset(gff, type == "mRNA") #keep only transcripts 

transcripts_gr <- makeGRangesFromDataFrame(transcripts, keep.extra.columns=TRUE) #extract length information

transcript_lengths <- width(transcripts_gr) #isolate length of each gene

seqnames<-transcripts_gr$ID #extract list of gene id 

lengths<-cbind(seqnames, transcript_lengths)

lengths<-as.data.frame(lengths) #convert to data frame
#remove unwanted character strings from the length dataframe
lengths <- lengths %>%
  mutate(seqnames = str_remove(seqnames, "-RA"))
#make seqnames lower case to make same as other data frames
lengths <- lengths %>%
  mutate(seqnames = tolower(seqnames))
```

Already has a lengths variable in Ahya_annot_seq_name_filtered. Now tr
```{r}
Ahya_annot_seq_name_filtered$length<-lengths$transcript_lengths[match(Ahya_annot_seq_name_filtered$SeqName, lengths$seqnames)]

which(is.na(Ahya_annot_seq_name_filtered$length)) #all genes have lengths 
```