---
title: "porites_DEG"
author: "LM Gorman, AS Huffmyer"
date: "2025-03-26"
output:
  html_document:
    toc: yes
    toc_depth: 6
    toc_float: yes
  pdf_document:
    keep_tex: yes
editor_options: 
  chunk_output_type: console
---

Set up workspace, set options, and load required packages.    
```{r}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries. 
```{r}
options(stringsAsFactors = FALSE)
#if ("tidyverse" %in% rownames(installed.packages()) == 'FALSE') install.packages('tidyverse') 
#if ("genefilter" %in% rownames(installed.packages()) == 'FALSE') install.packages('genefilter') 
#if ("RColorBrewer" %in% rownames(installed.packages()) == 'FALSE') install.packages('RColorBrewer') 
#if ("WGCNA" %in% rownames(installed.packages()) == 'FALSE') install.packages('WGCNA') 
#if ("flashClust" %in% rownames(installed.packages()) == 'FALSE') install.packages('flashClust') 
#if ("gridExtra" %in% rownames(installed.packages()) == 'FALSE') install.packages('gridExtra') 
#if ("ComplexHeatmap" %in% rownames(installed.packages()) == 'FALSE') BiocManager::install("ComplexHeatmap")
#if ("DESeq2" %in% rownames(installed.packages()) == 'FALSE') BiocManager::install("DESeq2")
#if ("goseq" %in% rownames(installed.packages()) == 'FALSE') BiocManager::install("goseq") 
#if ("dplyr" %in% rownames(installed.packages()) == 'FALSE') install.packages('dplyr') 
#if ("clusterProfiler" %in% rownames(installed.packages()) == 'FALSE') BiocManager::install("clusterProfiler") 
#if ("pheatmap" %in% rownames(installed.packages()) == 'FALSE') install.packages('pheatmap') 
#if ("magrittr" %in% rownames(installed.packages()) == 'FALSE') install.packages('magrittr') 
#if ("rtracklayer" %in% rownames(installed.packages()) == 'FALSE') BiocManager::install("rtracklayer")
#if ("GenomicRanges" %in% rownames(installed.packages()) == 'FALSE') BiocManager::install("GenomicRanges")
#if ("plyranges" %in% rownames(installed.packages()) == 'FALSE') BiocManager::install("plyranges")
#if ("GSEABase" %in% rownames(installed.packages()) == 'FALSE') BiocManager::install("GSEABase")
#if ("GOSim" %in% rownames(installed.packages()) == 'FALSE') BiocManager::install("GOSim")
#if ("stats" %in% rownames(installed.packages()) == 'FALSE') install.packages("stats")
#if ("ggdendro" %in% rownames(installed.packages()) == 'FALSE') BiocManager::install("ggdendro")
#if ("GO.db" %in% rownames(installed.packages()) == 'FALSE') BiocManager::install("GO.db")
#if ("rrvgo" %in% rownames(installed.packages()) == 'FALSE') BiocManager::install("rrvgo")
#if ("viridis" %in% rownames(installed.packages()) == 'FALSE') install.packages("viridis")
#if ("topGO" %in% rownames(installed.packages()) == 'FALSE') BiocManager::install("topGO")
#if ("biomaRt" %in% rownames(installed.packages()) == 'FALSE') BiocManager::install("biomaRt")
#if ("Rgraphviz" %in% rownames(installed.packages()) == 'FALSE') BiocManager::install("Rgraphviz")
#if ("EnhancedVolcano" %in% rownames(installed.packages()) == 'FALSE') BiocManager::install("EnhancedVolcano")
#if ("vegan" %in% rownames(installed.packages()) == 'FALSE') install.packages("vegan") 
#if ("factoextra" %in% rownames(installed.packages()) == 'FALSE') install.packages("factoextra") 

library("vegan")
library("factoextra")
library("viridis")
library("DESeq2")
library("tidyverse")
library("genefilter")
library("RColorBrewer")
library("WGCNA")
library("flashClust")
library("gridExtra")
library("ComplexHeatmap")
library("goseq")
library("dplyr")
library("clusterProfiler")
library("pheatmap")
library("magrittr")
library("rtracklayer")
library("GenomicRanges")
library("plyranges")
library("GSEABase")
library("GOSlim")
library("stats")
library("ggdendro")
library("GO.db")
library("rrvgo")
library("cowplot")
library("topGO")
library("biomaRt")
library("Rgraphviz")
library("EnhancedVolcano")
```

# Load and format data 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Load metadata
```{r}
metadata_por <- read.csv("data/metadata_porites.csv", header = TRUE, sep = ",")%>%dplyr::select(sample, eatenvscontrol, code)
metadata_por$code<-paste0(metadata_por$code)
head(metadata_por)
```

#Load Porites gene count matrix generated from stringtie
```{r}
gcount_por <- as.data.frame(read.csv("data/gene-count-matrix/porites_gene_count_matrix.csv", row.names="gene_id"), colClasses = double)
head(gcount_por)
#Check that there are no genes with 0 counts across all samples (remember only 16 columns since we are just looking at Porites). 
dim(gcount_por) 
   #[1] 40389    16
gcount_por<-gcount_por %>%
  mutate(Total = rowSums(.[, 1:16]))%>%
  filter(!Total==0)%>%
  dplyr::select(!Total)
dim(gcount_por)
  #[1] 35599    16
```

```{r}
#Create filter for the counts data
filt_por <- filterfun(pOverA(0.5,10))
gfilt_por <- genefilter(gcount_por, filt_por)
#identify genes to keep by count filter
gkeep_por <- gcount_por[gfilt_por,]
#identify gene lists
gn.keep_por <- rownames(gkeep_por)
#gene count data filtered in PoverA, P percent of the samples have counts over A
gcount_filt_por <- as.data.frame(gcount_por[which(rownames(gcount_por) %in% gn.keep_por),])

#How many rows do we have before and after filtering?
nrow(gcount_por) #Before
    #[1] 35599
nrow(gcount_filt_por) #After
    #[1] 16279
```

```{r}
# Function to extract 'R' followed by numbers from column names
extract_R_numbers_por <- function(col_names) {
  sub("R(\\d+).*", "\\1", col_names)
}

# Apply the function to all column names
new_col_names_por <- sapply(names(gcount_filt_por), extract_R_numbers_por)

# Rename the columns - do I need to do this?
# names(gcount_filt_por) <- paste0("R", new_col_names_por)
# Output the modified dataframe
names(gcount_filt_por)
length(names(gcount_filt_por))
    #[1] 16
```
Column names are now correct and there are 16 as expected.


```{r}
# Print the sample names in the metadata_por file. 
metadata_por$sample<-as.factor(metadata_por$sample)
metadata_por$eatenvscontrol<-as.factor(metadata_por$eatenvscontrol)
metadata_por$code<-as.factor(metadata_por$code)

# Set levels of factors. 
metadata_por$eatenvscontrol<-factor(metadata_por$eatenvscontrol, levels=c("eaten", "control"))
metadata_por$code<-factor(metadata_por$code, levels=c("Por_eaten", "Por_control"))

#Make sure the metadata and the columns in the gene count matrix are all the same.  
metadata_por$sample
colnames(gcount_filt_por)
list_por<-colnames(gcount_filt_por)
list_por<-as.factor(list_por)

#Make sure metadata$sample and column names are identical
metadata_por$sample <- tolower(trimws(metadata_por$sample))
colnames(gcount_filt_por) <- tolower(trimws(colnames(gcount_filt_por)))
identical(metadata_por$sample, colnames(gcount_filt_por))
    #[1] TRUE

## Re-order the levels - havent included the second part as it was messing my code up!
metadata_por$sample<-as.factor(metadata_por$sample)
##metadata_por$sample <- factor(as.character(metadata_por$sample), levels=list)

# Re-order the data.frame
metadata_por_ordered <- metadata_por[order(metadata_por$sample),]
metadata_por_ordered$sample

head(metadata_por_ordered)

metadata_por_ordered$sample
colnames(gcount_filt_por)
head(gcount_filt_por)

gcount_filt_por <- gcount_filt_por[, order(colnames(gcount_filt_por))]

# Remove all spaces from the 'sample' column
metadata_por_ordered$sample <- gsub("\\s+", "", metadata_por_ordered$sample)
# Remove all spaces from the column names of 'gcount_filt_por'
colnames(gcount_filt_por) <- gsub("\\s+", "", colnames(gcount_filt_por))
#Check names are the same
identical(metadata_por_ordered$sample, colnames(gcount_filt_por))
    #[1] TRUE

## Check unique sample names from metadata
unique(metadata_por_ordered$sample)
# Check unique column names from gcount_filt_por
unique(colnames(gcount_filt_por))
# Clean up metadata sample names
metadata_por_ordered$sample <- tolower(trimws(metadata_por_ordered$sample))

# Clean up gcount column names
colnames(gcount_filt_por) <- tolower(trimws(colnames(gcount_filt_por)))
identical(metadata_por_ordered$sample, colnames(gcount_filt_por))
    #[1] TRUE
#Make sure that the reference level for estimating effect size is control
metadata_por$eatenvscontrol <- relevel(metadata_por$eatenvscontrol, ref = "control")
```

```{r}
gdds_por <- DESeqDataSetFromMatrix(countData = gcount_filt_por,
                               colData = metadata_por,
                               design = ~eatenvscontrol)

SF.gdds_por <- estimateSizeFactors(gdds_por) #estimate size factors to determine if we can use vst  to transform our data. Size factors should be less than 4 for us to use vst
print(sizeFactors(SF.gdds_por)) #View size factors

all(sizeFactors(SF.gdds_por)) < 4
    #[1] TRUE (but look like we have 11 and 8 when we print the factors - ask Ariana)

#All size factors are less than 4, so we can use VST transformation.  
gvst_por <- vst(gdds_por, blind=FALSE) #apply a variance stabilizing transforamtion to minimize effects of small counts and normalize wrt library size
head(assay(gvst_por), 3) #view transformed gene count data for the first three genes in the dataset. 

dim(gvst_por)
    #[1] 16279    16
```


#Plot a heatmap to sample distances
```{r}
gsampleDists_por <- dist(t(assay(gvst_por))) #calculate distance matix
gsampleDistMatrix_por <- as.matrix(gsampleDists_por) #distance matrix
rownames(gsampleDistMatrix_por) <- colnames(gvst_por) #assign row names
colnames(gsampleDistMatrix_por) <- NULL #assign col names
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255) #assign colors
save_pheatmap_pdf <- function(x, filename, width=7, height=7) {
  stopifnot(!missing(x))
  stopifnot(!missing(filename))
  pdf(filename, width=width, height=height)
  grid::grid.newpage()
  grid::grid.draw(x$gtable)
  dev.off()
}

pht_por<-pheatmap(gsampleDistMatrix_por, #plot matrix
              clustering_distance_rows=gsampleDists_por, #cluster rows
              clustering_distance_cols=gsampleDists_por, #cluster columns
              col=colors) #set colors

save_pheatmap_pdf(pht, "figures/porites/sample_distances_por.pdf")
dev.off()
```

#Conduct PERMANOVA and PCA for all genes
```{r}
#Export data for PERMANOVA test.  
test_por<-t(assay(gvst_por)) #export as matrix
test_por<-as.data.frame(test_por)

#add category columns
test_por$sample<-rownames(test_por)
test_por$eatenvscontrol<-metadata_por$eatenvscontrol[match(test_por$sample, metadata_por$sample)]

#Build PERMANOVA model for all genes in the dataset.  
dim(test_por)
    #[1]    16 16281 (remember two of the values are for the right measurement are just names so actually = 16279)
scaled_test_por <-prcomp(test_por[c(1:16279)], scale=TRUE, center=TRUE)
fviz_eig(scaled_test_por)

# scale data
vegan_por <- scale(test_por[c(1:16279)])

# PerMANOVA 
permanova_por<-adonis2(vegan_por ~ eatenvscontrol, data = test_por, method='eu')
permanova_por
```
Permutation test for adonis under reduced model
Permutation: free
Number of permutations: 999
adonis2(formula = vegan_por ~ eatenvscontrol, data = test_por, method = "eu")
Df SumOfSqs     R2      F Pr(>F)  
Model     1    38800 0.1589 2.6448  0.064 .
Residual 14   205385 0.8411                
Total    15   244185 1.0000                
---
  Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1


#Plot a PCA of samples by eatenvscontrol phenotype for all genes. 
```{r}
gPCAdata_por <- plotPCA(gvst_por, intgroup = c("eatenvscontrol"), returnData=TRUE)
percentVar_por <- round(100*attr(gPCAdata_por, "percentVar")) #plot PCA of samples with all data

allgenesfilt_PCA_por <- ggplot(gPCAdata_por, aes(PC1, PC2, color=eatenvscontrol)) + 
  geom_point(size=3) + 
  ggtitle(expression(bold("All genes (16279 genes)")))+
  geom_text(aes(x = PC1, y = PC2, label = name), vjust = -0.5) +
  xlab(paste0("PC1: ",percentVar_por[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar_por[2],"% variance")) +
  scale_color_manual(name="Eaten vs Control", values=c("control"="gray", "eaten"="orange"))+
  geom_text(x=0, y=10, label="p[eatenvscontrol]=0.065", color="black")+
  theme_classic() + #Set background color
  theme(panel.border = element_blank(), # Set border
        #panel.grid.major = element_blank(), #Set major gridlines 
        #panel.grid.minor = element_blank(), #Set minor gridlines
        axis.line = element_line(colour = "black"), #Set axes color
        plot.background=element_blank()); allgenesfilt_PCA_por

ggsave("figures/porites/pca_por.pdf", allgenesfilt_PCA_por, width=8, height=4)
ggsave("figures/porites/pca_por.jpeg", allgenesfilt_PCA_por, width=8, height=4)

#Add a plot with elipses.  
allgenes_ellipse_por<-allgenesfilt_PCA_por + stat_ellipse();allgenes_ellipse_por
ggsave("figures/porites/pca_ellipse_por.pdf", allgenes_ellipse_por, width=9, height=5)
ggsave("figures/porites/pca_ellipse_por.jpeg", allgenes_ellipse_por, width=9, height=5)
```
Remove left cluster as it is clusering only based on read depth (225R, 34R, 43R, 16R)

#Remove 225R, 34R, 43R and 16R from metadata 
```{r}
metadata_por2 <- metadata_por[c(-1, -4, -10, -11), ]
```

#Remove 225R, 34R, 43R and 16R from gene count matrix. 
```{r}
#Remove 497R
gcount_por2 <- gcount_por[,c(-1, -4, -10, -11)]
```

#Re-run code with removed samples (225R, 34R, 43R and 16R)
```{r}
dim(gcount_por2) 
   #[1] 
gcount_por2<-gcount_por2 %>%
  mutate(Total = rowSums(.[, 1:12]))%>%
  filter(!Total==0)%>%
  dplyr::select(!Total)
dim(gcount_por2)
  #[1] 35577    12
```

##pOver A 4/12=0.33
```{r}
#Create filter for the counts data
filt_por2 <- filterfun(pOverA(0.33,10))
gfilt_por2 <- genefilter(gcount_por2, filt_por2)
#identify genes to keep by count filter
gkeep_por2 <- gcount_por2[gfilt_por2,]
#identify gene lists
gn.keep_por2 <- rownames(gkeep_por2)
#gene count data filtered in PoverA, P percent of the samples have counts over A
gcount_filt_por2 <- as.data.frame(gcount_por2[which(rownames(gcount_por2) %in% gn.keep_por2),])

#How many rows do we have before and after filtering?
nrow(gcount_por2) #Before
    #[1] 35577
nrow(gcount_filt_por2) #After
    #[1] 24301
```
Now has around 8,000 more genes after removing samples (24,000 versus 16,000)

```{r}
# Function to extract 'R' followed by numbers from column names
extract_R_numbers_por2 <- function(col_names) {
  sub("R(\\d+).*", "\\1", col_names)
}

# Apply the function to all column names
new_col_names_por2 <- sapply(names(gcount_filt_por2), extract_R_numbers_por2)

# Rename the columns - do I need to do this?
# names(gcount_filt_por) <- paste0("R", new_col_names_por)
# Output the modified dataframe
names(gcount_filt_por2)
length(names(gcount_filt_por2))
    #[1] 12
```
Column names are now correct and there are 12 as expected.


```{r}
# Print the sample names in the metadata_por file. 
metadata_por2$sample<-as.factor(metadata_por2$sample)
metadata_por2$eatenvscontrol<-as.factor(metadata_por2$eatenvscontrol)
metadata_por2$code<-as.factor(metadata_por2$code)

# Set levels of factors. 
metadata_por2$eatenvscontrol<-factor(metadata_por2$eatenvscontrol, levels=c("eaten", "control"))
metadata_por2$code<-factor(metadata_por2$code, levels=c("Por_eaten", "Por_control"))

#Make sure the metadata and the columns in the gene count matrix are all the same.  
metadata_por2$sample
colnames(gcount_filt_por2)
list_por2<-colnames(gcount_filt_por2)
list_por2<-as.factor(list_por2)

#Make sure metadata$sample and column names are identical
metadata_por2$sample <- tolower(trimws(metadata_por2$sample))
colnames(gcount_filt_por2) <- tolower(trimws(colnames(gcount_filt_por2)))
identical(metadata_por2$sample, colnames(gcount_filt_por2))
    #[1] TRUE

## Re-order the levels - havent included the second part as it was messing my code up!
metadata_por2$sample<-as.factor(metadata_por2$sample)
##metadata_por$sample <- factor(as.character(metadata_por$sample), levels=list)

# Re-order the data.frame
metadata_por2_ordered <- metadata_por2[order(metadata_por2$sample),]
metadata_por2_ordered$sample

head(metadata_por2_ordered)

metadata_por2_ordered$sample
colnames(gcount_filt_por2)
head(gcount_filt_por2)

gcount_filt_por2 <- gcount_filt_por2[, order(colnames(gcount_filt_por2))]

# Remove all spaces from the 'sample' column
metadata_por2_ordered$sample <- gsub("\\s+", "", metadata_por2_ordered$sample)
# Remove all spaces from the column names of 'gcount_filt_por'
colnames(gcount_filt_por2) <- gsub("\\s+", "", colnames(gcount_filt_por2))
#Check names are the same
identical(metadata_por2_ordered$sample, colnames(gcount_filt_por2))
    #[1] TRUE

## Check unique sample names from metadata
unique(metadata_por2_ordered$sample)
# Check unique column names from gcount_filt_por
unique(colnames(gcount_filt_por2))
# Clean up metadata sample names
metadata_por2_ordered$sample <- tolower(trimws(metadata_por2_ordered$sample))

# Clean up gcount column names
colnames(gcount_filt_por2) <- tolower(trimws(colnames(gcount_filt_por2)))
identical(metadata_por2_ordered$sample, colnames(gcount_filt_por2))
    #[1] TRUE
#Make sure that the reference level for estimating effect size is control
metadata_por2$eatenvscontrol <- relevel(metadata_por2$eatenvscontrol, ref = "control")
```

```{r}
gdds_por2 <- DESeqDataSetFromMatrix(countData = gcount_filt_por2,
                               colData = metadata_por2,
                               design = ~eatenvscontrol)

SF.gdds_por2 <- estimateSizeFactors(gdds_por2) #estimate size factors to determine if we can use vst  to transform our data. Size factors should be less than 4 for us to use vst
print(sizeFactors(SF.gdds_por2)) #View size factors

all(sizeFactors(SF.gdds_por2)) < 4
    #[1] TRUE (but look like we have 11 and 8 when we print the factors - ask Ariana)

#All size factors are less than 4, so we can use VST transformation.  
gvst_por2 <- vst(gdds_por2, blind=FALSE) #apply a variance stabilizing transforamtion to minimize effects of small counts and normalize wrt library size
head(assay(gvst_por2), 3) #view transformed gene count data for the first three genes in the dataset. 

dim(gvst_por2)
    #[1] 24301    12
```


#Plot a heatmap to sample distances
```{r}
gsampleDists_por2 <- dist(t(assay(gvst_por2))) #calculate distance matix
gsampleDistMatrix_por2 <- as.matrix(gsampleDists_por2) #distance matrix
rownames(gsampleDistMatrix_por2) <- colnames(gvst_por2) #assign row names
colnames(gsampleDistMatrix_por2) <- NULL #assign col names
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255) #assign colors
save_pheatmap_pdf <- function(x, filename, width=7, height=7) {
  stopifnot(!missing(x))
  stopifnot(!missing(filename))
  pdf(filename, width=width, height=height)
  grid::grid.newpage()
  grid::grid.draw(x$gtable)
  dev.off()
}

pht_por2<-pheatmap(gsampleDistMatrix_por2, #plot matrix
              clustering_distance_rows=gsampleDists_por2, #cluster rows
              clustering_distance_cols=gsampleDists_por2, #cluster columns
              col=colors) #set colors

save_pheatmap_pdf(pht_por2, "figures/porites/sample_distances_por2.pdf")
dev.off()
```
Diagnol = great


#Conduct PERMANOVA and PCA for all genes
```{r}
#Export data for PERMANOVA test.  
test_por2<-t(assay(gvst_por2)) #export as matrix
test_por2<-as.data.frame(test_por2)

#add category columns
test_por2$sample<-rownames(test_por2)
test_por2$eatenvscontrol<-metadata_por2$eatenvscontrol[match(test_por2$sample, metadata_por2$sample)]

#Build PERMANOVA model for all genes in the dataset.  
dim(test_por2)
    #[1]    12 24303 (remember two of the values are for the right measurement are just names so actually = 24301)
scaled_test_por2 <-prcomp(test_por2[c(1:24301)], scale=TRUE, center=TRUE)
fviz_eig(scaled_test_por2)

# scale data
vegan_por2 <- scale(test_por2[c(1:24301)])

# PerMANOVA 
permanova_por2<-adonis2(vegan_por2 ~ eatenvscontrol, data = test_por2, method='eu')
permanova_por2
```
Permutation test for adonis under reduced model
Permutation: free
Number of permutations: 999
adonis2(formula = vegan_por ~ eatenvscontrol, data = test_por, method = "eu")
 Df SumOfSqs      R2      F Pr(>F)
Model     1    21700 0.08118 0.8835  0.559
Residual 10   245611 0.91882              
Total    11   267311 1.00000                   
---
  Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1


#Plot a PCA of samples by eatenvscontrol phenotype for all genes. 
```{r}
gPCAdata_por2 <- plotPCA(gvst_por2, intgroup = c("eatenvscontrol"), returnData=TRUE)
percentVar_por2 <- round(100*attr(gPCAdata_por2, "percentVar")) #plot PCA of samples with all data

allgenesfilt_PCA_por2 <- ggplot(gPCAdata_por2, aes(PC1, PC2, color=eatenvscontrol)) + 
  geom_point(size=3) + 
  ggtitle(expression(bold("All genes (24301 genes)")))+
  geom_text(aes(x = PC1, y = PC2, label = name), vjust = -0.5) +
  xlab(paste0("PC1: ",percentVar_por2[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar_por[2],"% variance")) +
  scale_color_manual(name="Eaten vs Control", values=c("control"="gray", "eaten"="orange"))+
  geom_text(x=0, y=10, label="p[eatenvscontrol]=0.559", color="black")+
  theme_classic() + #Set background color
  theme(panel.border = element_blank(), # Set border
        #panel.grid.major = element_blank(), #Set major gridlines 
        #panel.grid.minor = element_blank(), #Set minor gridlines
        axis.line = element_line(colour = "black"), #Set axes color
        plot.background=element_blank()); allgenesfilt_PCA_por2

ggsave("figures/porites/pca_por2.pdf", allgenesfilt_PCA_por2, width=8, height=4)
ggsave("figures/porites/pca_por2.jpeg", allgenesfilt_PCA_por2, width=8, height=4)

#Add a plot with elipses.  
allgenes_ellipse_por2<-allgenesfilt_PCA_por2 + stat_ellipse();allgenes_ellipse_por2
ggsave("figures/porites/pca_ellipse_por2.pdf", allgenes_ellipse_por2, width=9, height=5)
ggsave("figures/porites/pca_ellipse_por2.jpeg", allgenes_ellipse_por2, width=9, height=5)
```


#Run DEG analysis of eatenvscontrol   
Use the likelihood ratio approach based on guidance from DESeq2: "DESeq2 offers two kinds of hypothesis tests: the Wald test, where we use the estimated standard error of a log2 fold change to test if it is equal to zero, and the likelihood ratio test (LRT). The LRT examines two models for the counts, a full model with a certain number of terms and a reduced model, in which some of the terms of the full model are removed. The test determines if the increased likelihood of the data using the extra terms in the full model is more than expected if those extra terms are truly zero.
The LRT is therefore useful for testing multiple terms at once, for example testing 3 or more levels of a factor at once, or all interactions between two variables. The LRT for count data is conceptually similar to an analysis of variance (ANOVA) calculation in linear regression, except that in the case of the Negative Binomial GLM, we use an analysis of deviance (ANODEV), where the deviance captures the difference in likelihood between a full and a reduced model." 


```{r} 
#Run Wald test.  
DEG_T_por2 <- DESeq(gdds_por2, test="Wald")
#View results by eatenvscontrol effects. Create dataframe for each comparison and add a column for the specific contrast.  
resultsNames(DEG_T_por2)
#this shows the comparisons made (shown similar to lm output in terms of interactions)
eatenvscontrol_por2<-results(DEG_T_por2, contrast=c("eatenvscontrol","control","eaten"))
eatenvscontrol_por2 <- as.data.frame(subset(eatenvscontrol_por2, padj<0.01))
eatenvscontrol_por2 <- as.data.frame(subset(eatenvscontrol_por2, abs(log2FoldChange)>1.5)
eatenvscontrol_por2$contrast <- c("eatenvscontrol_por")
eatenvscontrol_por2$gene <- rownames(eatenvscontrol_por2)

#Try extracting column names
# Subset row names from colData in DEG_T_por2 - need help to do this to correct genes
colData_rownames <- rownames(DEG_T_por2@colData)
#Not working
eatenvscontrol_por2$sample <- ifelse(
  abs(DEG_T_por2$log2FoldChange) > 1.5 & DEG_T_por2$padj < 0.01, 
  DEG_T_por2$colData, 
  NA
)

# Check the result
head(eatenvscontrol_por2)

rownames(eatenvscontrol_por2) <- NULL
dim(eatenvscontrol_por2)
# [1] 16   8
```
16 DEGs between eaten and control Porites colonies.

Combine this list of DEGs into one dataframe. 
```{r}
DEG_eatenvscontrol_por2<-rbind(eatenvscontrol_por2)
dim(DEG_eatenvscontrol_por2)
length(unique(DEG_eatenvscontrol_por2$gene))

#There are 16 total DEGs.
#Get a VST normalized gene count matrix for these DEGs. 
DEG_eatenvscontrol_vst_por2 <- gdds_por2[unique(DEG_eatenvscontrol_por2$gene)]
dim(DEG_eatenvscontrol_vst_por2)

# Now apply variance stabilizing transformation
DEG_eatenvscontrol_vst_por2 <- varianceStabilizingTransformation(gcount_filt_por2)

DEG_eatenvscontrol_vst_por2 <- varianceStabilizingTransformation(DEG_eatenvscontrol_vst_por2) 
```

#Plot a heatmap for eatenvscontrol DEGs.
```{r}
#Set themes and metadata.  
colnames(colData(gdds_por2))
legend_names_col_por <- colnames(assay(DEG_eatenvscontrol_vst_por2))
df_por2 <- as.data.frame(colData(gdds_por2))
colnames(df_por2) <- tolower(trimws(colnames(df_por2)))
df_por2$eatenvscontrol <- factor(as.character(df_por2$eatenvscontrol), levels=c("eaten", "control"))
df_por2$eatenvscontrol <- relevel(df_por2$eatenvscontrol, ref = "control")
 
df_por2<-df_por2%>%
  arrange(eatenvscontrol)

ann_colors_por = list(eatenvscontrol = c("control" = "lightblue", "eaten"= "red2"))


DEG_por2_eatenvscontrol_heatmap <- pheatmap(assay(DEG_eatenvscontrol_vst_por2), cluster_rows=TRUE, show_rownames=FALSE, color=inferno(10), show_colnames=TRUE, fontsize_row=3, scale="row", cluster_cols=TRUE, annotation_col=df_por2, annotation_colors = ann_colors_por, labels_col=legend_names_col_por)

save_pheatmap_pdf(DEG_por2_eatenvscontrol_heatmap, "figures/porites/DEG_por2_eatenvscontrol_heatmap.pdf", width=15, height=20)
dev.off()
```