---
title: "porites_functional_enrichment"
author: "LM Gorman, AS Huffmyer"
date: "2025-03-28"
output:
  html_document:
    toc: yes
    toc_depth: 6
    toc_float: yes
  pdf_document:
    keep_tex: yes
editor_options: 
  chunk_output_type: console
---

# Set up 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# The following setting is important, do not omit.
options(stringsAsFactors = FALSE) #Set Strings to character
```

```{r}
if ("tidyverse" %in% rownames(installed.packages()) == 'FALSE') install.packages('tidyverse') 
if ("genefilter" %in% rownames(installed.packages()) == 'FALSE') install.packages('genefilter') 
if ("DESeq2" %in% rownames(installed.packages()) == 'FALSE') install.packages('DESeq2') 
if ("RColorBrewer" %in% rownames(installed.packages()) == 'FALSE') install.packages('RColorBrewer') 
if ("WGCNA" %in% rownames(installed.packages()) == 'FALSE') install.packages('WGCNA') 
if ("flashClust" %in% rownames(installed.packages()) == 'FALSE') install.packages('flashClust') 
if ("gridExtra" %in% rownames(installed.packages()) == 'FALSE') install.packages('gridExtra') 
if ("ComplexHeatmap" %in% rownames(installed.packages()) == 'FALSE') install.packages('ComplexHeatmap') 
if ("goseq" %in% rownames(installed.packages()) == 'FALSE') install.packages('goseq') 
if ("dplyr" %in% rownames(installed.packages()) == 'FALSE') install.packages('dplyr') 
if ("clusterProfiler" %in% rownames(installed.packages()) == 'FALSE') install.packages('clusterProfiler') 
if ("pheatmap" %in% rownames(installed.packages()) == 'FALSE') install.packages('pheatmap') 
if ("magrittr" %in% rownames(installed.packages()) == 'FALSE') install.packages('magrittr') 
if ("rtracklayer" %in% rownames(installed.packages()) == 'FALSE') BiocManager::install("rtracklayer")
if ("GenomicRanges" %in% rownames(installed.packages()) == 'FALSE') BiocManager::install("GenomicRanges")
if ("plyranges" %in% rownames(installed.packages()) == 'FALSE') BiocManager::install("plyranges")
if ("GSEABase" %in% rownames(installed.packages()) == 'FALSE') BiocManager::install("GSEABase")
#if ("GOSim" %in% rownames(installed.packages()) == 'FALSE') BiocManager::install("GOSim")
if ("stats" %in% rownames(installed.packages()) == 'FALSE') install.packages("stats")
if ("ggdendro" %in% rownames(installed.packages()) == 'FALSE') BiocManager::install("ggdendro")
if ("GO.db" %in% rownames(installed.packages()) == 'FALSE') BiocManager::install("GO.db")
if ("rrvgo" %in% rownames(installed.packages()) == 'FALSE') BiocManager::install("rrvgo")

#BiocManager::install("org.Ce.eg.db", force=TRUE) #install if needed 

library("tidyverse")
library("genefilter")
library("DESeq2")
library("RColorBrewer")
library("WGCNA")
library("flashClust")
library("gridExtra")
library("ComplexHeatmap")
library("goseq")
library("dplyr")
library("clusterProfiler")
library("pheatmap")
library("magrittr")
library("rtracklayer")
library("GenomicRanges")
library("plyranges")
library("GSEABase")
#library("GOSim")
library("stats")
library("ggdendro")
library("GO.db")
library("rrvgo")
library("cowplot")
library("ape")
library("topGO")
```
# Read in data files 

1. Read in file with vst transformed counts of all genes detected and kept after filtering. 
```{r}
#Need to work out which file this is from porites_DEG script
all_genes_por<-read.csv("data/all_genes_por.csv")
colnames(all_genes_por)[1] <- "gene_id"
transposed_all_genes_por <- t(all_genes_por)
print(transposed_all_genes_por)
transposed_all_genes_por <- as.data.frame(t(all_genes_por))
# Set proper column names using row names (which were originally column headers)
colnames(transposed_all_genes_por) <- transposed_all_genes_por[1, ]  # First row as column names
transposed_all_genes_por <- transposed_all_genes_por[-1, ] 
```
2. Read in file of DEG's between eaten and control
```{r}
por_eatenvscontrol_DEG_list<-read.csv("data/DEG_por.csv")
por_eatenvscontrol_DEG_list$contrast<-factor(por_eatenvscontrol_DEG_list$contrast, levels=c("eaten", "control"))
por_eatenvscontrol_DEG_list<- por_eatenvscontrol_DEG_list %>% 
  rename(GeneID = gene)
```

# Read in annotation files 
```{r, Lucys functional annotation from omicsbox}
Por_annot <- read.delim("data/Pevermanni_metadata.txt",  header=TRUE, na.strings = "---NA---")
dim(Por_annot)
probes <- all_genes_por$gene_id
filtered_genes_por <- Por_annot[Por_annot$Sequence.Name %in% probes,]
filtered_genes_por<- filtered_genes_por %>% 
  rename(GeneID = Sequence.Name)
write.csv(filtered_genes_por,file='data/Por_annot_filtered.csv') #22,214 genes
```

# TopGO 

In topGO we need: 

- Ontology: character string specifying the ontology of interest (BP, MF or CC).
- allGenes: named vector of type numeric or factor. The names attribute contains the genes identifiers. The genes listed in this object define the gene universe.
- nodeSize: an integer larger or equal to 1. This parameter is used to prune the GO hierarchy from the terms which have less than nodeSize annotated genes (after the true path rule is applied).
- annotationFun: function which maps genes identifiers to GO terms. There are a couple of annotation function included in the package trying to address the user’s needs. The annotation functions take three arguments. One of those arguments is specifying where the mappings can be found, and needs to be provided by the user. annFUN.gene2GO this function is used when the annotations are provided as a gene-to-GOs mapping.

Background/GO information: 
```{r}
geneID2GO_por<-filtered_genes_por%>%dplyr::select(GeneID, Annotation.GO.ID)
#geneID2GO<-Ahya_annot_seq_name_filtered_lucy 
geneID2GO_por <- as.data.frame(geneID2GO_por)
background_por <- geneID2GO_por%>%pull("GeneID")

#remove genes with NA in go term and separate rows based on GO terms 
geneID2GO_por<-geneID2GO_por%>%
  filter(!(Annotation.GO.ID == ""))%>%
  separate_rows(Annotation.GO.ID, sep = ";")
#creates list with each GO term appended to the geneID
geneID2GO_por <- split(as.character(geneID2GO_por$Annotation.GO.ID), geneID2GO_por$GeneID)
```

Run TopGO

```{r}
por_test_set<-por_eatenvscontrol_DEG_list%>%pull(GeneID)
# 1. Make sure background is a character vector
background_por <- as.character(background_por)
# 2. Make sure acr_test_set is character too
por_test_set <- as.character(por_test_set)
# 3. Build geneList_por as a named factor
geneList_por <- factor(background_por %in% por_test_set, levels = c(FALSE, TRUE))
# 4. Check structure
str(geneList_por)
#Make sure the geneList_por is an integer NOT a logical
geneList_por <- as.integer(background_por %in% por_test_set)
names(geneList_por) <- background_por
table(geneList_por) 

str(geneID2GO_por, max.level = 1)
head(geneID2GO_por, 3)
sum(names(geneList_por) %in% names(geneID2GO_por))

GOdata_por <- new("topGOdata",
              description = "por_DEGs",
              ontology = "BP",
              allGenes = geneList_por,
              geneSelectionFun = function(x) x == 1,
              nodeSize = 5,
              gene2GO = geneID2GO_por, annot = annFUN.gene2GO) #Gene

GOdata_por
numSigGenes(GOdata_por) #6 sig genes
sum(feasible(GOdata_por)) #13991
```

Run fisher test. 
```{r}
resultFisher_por <- runTest(GOdata_por, algorithm = "classic", statistic = "fisher")
resultFisher_por
hist(score(resultFisher_por), 50, xlab = "p-values")

#33 terms with p<0.01

resultWeight_por <- runTest(GOdata_por, algorithm = "weight01", statistic = "fisher")
resultWeight_por
hist(score(resultWeight_por), 50, xlab = "p-values")
#9 terms with p<0.01

# p < 0.05 for weighted fisher
#Generate GO term table from resultWeight
allGO_por <- GenTable(GOdata_por,
                  weight01Fisher = resultWeight_por,
                  orderBy = "weight01Fisher",
                  topNodes = 1000)  # get many to filter later

# Convert p-values to numeric
allGO_por$weight01Fisher <- as.numeric(allGO_por$weight01Fisher)

# Filter for p < 0.05
signifGO_por <- subset(allGO_por, weight01Fisher < 0.05)

# See how many significant terms
nrow(signifGO_por) #16 GO terms
head(signifGO_por)

#p<0.05 for classic fisher
classicGO_por <- GenTable(GOdata_por,
                      classicFisher = resultFisher_por,
                      orderBy = "classicFisher",
                      topNodes = 1000)

classicGO_por$classicFisher <- as.numeric(classicGO_por$classicFisher)
signifClassic_por <- subset(classicGO_por, classicFisher < 0.05)

nrow(signifClassic_por)#126 GO terms
head(signifClassic_por)
```

Which test should be used? 

```{r}
allRes_por <- GenTable(GOdata_por, classicFisher = resultFisher_por, weightFisher = resultWeight_por, orderBy = "weightFisher", ranksOf = "classicFisher", topNodes = length(score(resultFisher_por)))

# Convert character columns to numeric
allRes_por$classicFisher <- as.numeric(allRes_por$classicFisher)
allRes_por$weightFisher <- as.numeric(allRes_por$weightFisher)

#adjust p-values 
allRes_por$bh_adjust <- p.adjust(allRes_por$classicFisher, method="BH") #add adjusted p-values

#adjust p-values 
allRes_por$bh_adjust_weight <- p.adjust(allRes_por$weightFisher, method="BH") #add adjusted p-values
```

```{r}
showSigOfNodes(GOdata_por, score(resultFisher_por), firstSigNodes = 5, useInfo = 'def')
dev.off()

showSigOfNodes(GOdata_por, score(resultWeight_por), firstSigNodes = 5, useInfo = 'def')
dev.off()
```

The subgraph induced by the top 5 GO terms identified by the classic algorithm for scoring GO terms for enrichment. Boxes indicate the 5 most significant terms. Box color represents the relative significance, ranging from dark red (most significant) to light yellow (least significant). Black arrows indicate is-a relationships and red arrows part-of relationships.

Proceed using weighted fisher test results.  
```{r}
p05_res_por <- allRes_por %>% dplyr::filter(weightFisher < 0.05)
  
#Reduce/collapse GO term set with the rrvgo package 
simMatrix_por <- calculateSimMatrix(p05_res_por$GO.ID,
                                orgdb="org.Ce.eg.db", #c. elegans database
                                ont="BP",
                                method="Rel")
```

```{r}
 #calculate similarity 
scores_por <- setNames(-log(as.numeric(p05_res_por$weightFisher)), p05_res_por$GO.ID)
reducedTerms_por <- reduceSimMatrix(simMatrix_por,
                                scores_por,
                                threshold=0.7,
                                orgdb="org.Ce.eg.db")
dim(reducedTerms_por)# 12 10
```

```{r}
#keep only the goterms from the reduced list
p05_res_reduced_por <- p05_res_por %>%
  filter(GO.ID %in% reducedTerms_por$go)

#add in parent terms to list of go terms 
p05_res_reduced_por$ParentTerm <- reducedTerms_por$parentTerm[match(p05_res_reduced_por$GO.ID, reducedTerms_por$go)]

length(unique(p05_res_reduced_por$ParentTerm))#7

p05_res_reduced_por <- p05_res_reduced_por %>% group_by(ParentTerm) %>% mutate("N_in_Parent" = dplyr::n()) %>% ungroup()
length(unique(p05_res_reduced_por$ParentTerm)) #7
```

```{r}
write.csv(p05_res_reduced_por, "output/por_parent_topGO_results.csv")
```

```{r}
parent_enrich_plot1_por <- ggplot(p05_res_reduced_por, aes(x=weightFisher,y=reorder(Term,weightFisher)))+
  facet_grid(ParentTerm ~ ., scales="free", labeller = label_wrap_gen(width = 10, multi_line = TRUE))+
  geom_point(size=2, color="black")+
  geom_segment(aes(x=0, xend=weightFisher, y=Term, yend=Term)) +
  #geom_hline(yintercept = 0, linetype="solid", color = 'black', linewidth=0.5, show.legend = TRUE)+
  #scale_y_continuous(limits=c(0,40))+
   #scale_x_discrete(labels = label_wrap(30)) +
  #scale_fill_gradientn(colours=c("white","#fddbc7","#f4a582","#d6604d","#b2182b"), na.value = "grey98",limits = c(0, 100)) +
  #labs(title = "Enriched Biological Process Parent GO Terms, Brown Module", x="Parent Term", y="Number of Significant (p < 0.01) GO Terms in Module", colour="p-value") +
  theme_classic() + 
  theme(axis.text.y = element_text(size = 8), 
              axis.title = element_text(size = 8), 
              plot.title = element_text(hjust = 0.5, size = 8));parent_enrich_plot1_por

ggsave("figures/porites/parent_enrich_plot1_por.pdf", plot = parent_enrich_plot1_por, width = 8, height = 15, units = "in")
dev.off()  # close and save

```

```{r}
p05_res_reduced_por$GeneRatio <- p05_res_reduced_por$Significant / p05_res_reduced_por$Annotated
parent_enrich_plot3_por <- ggplot(p05_res_reduced_por,
       aes(x = GeneRatio,
           y = reorder(Term, GeneRatio),
           color = -log10(weightFisher))) +
  geom_point(size=1) +
  geom_segment(aes(x=0, xend=GeneRatio, y=Term, yend=Term), color="grey70") +
  scale_color_gradient(low="lightblue", high="darkblue", name="-log10(weightFisher)") +
  theme_classic(base_size=10) +
  labs(x="Gene Ratio", y="GO Term") +
  theme(axis.text.y = element_text(size=8),
        axis.title = element_text(size=9),
        plot.title = element_text(hjust=0.5, size=10));parent_enrich_plot3_por
ggsave(parent_enrich_plot3_por, filename="figures/porites/por_parent_enrichment_individual_terms.png", width=10, height=11)

#summarize by parent term
parent_summary_por <- p05_res_reduced_por %>%
  group_by(ParentTerm) %>%
  summarise(
    maxLogP = max(-log10(weightFisher)),
    meanGeneRatio = mean(Significant/Annotated),
    nTerms = length(maxLogP)
  ) %>%
  arrange(desc(maxLogP))

parent_enrich_plot4_por<-ggplot(parent_summary_por, aes(x = meanGeneRatio, y = reorder(ParentTerm, meanGeneRatio), color = maxLogP, size = nTerms)) +
  geom_point(size=2) +
  scale_color_gradient(low = "lightblue", high = "darkblue", name="Max -log10(weightFisher)") +
  scale_size_continuous(name="Number of Terms") +
  theme_classic(base_size=10) +
  labs(x="Mean Gene Ratio", y="Parent GO Term");parent_enrich_plot4_por
ggsave(parent_enrich_plot4_por, filename="figures/porites/por_parent_enrichment_parent_terms.png", width=10, height=7)
```

# What about classic fisher results...

Proceed using classic fisher test results.  
```{r}
p05_clas_res_por <- allRes_por %>% dplyr::filter(classicFisher < 0.05)
  
#Reduce/collapse GO term set with the rrvgo package 
simMatrix_por_class <- calculateSimMatrix(p05_clas_res_por$GO.ID,
                                orgdb="org.Ce.eg.db", #c. elegans database
                                ont="BP",
                                method="Rel")
```

```{r}
 #calculate similarity 
scores_por_class <- setNames(-log(as.numeric(p05_clas_res_por$classicFisher)), p05_clas_res_por$GO.ID)
reducedTerms_por_class <- reduceSimMatrix(simMatrix_por_class,
                                scores_por_class,
                                threshold=0.7,
                                orgdb="org.Ce.eg.db")
dim(reducedTerms_por_class)# 74 10
```

```{r}
#keep only the goterms from the reduced list
p05_clas_res_reduced_por <- p05_clas_res_por %>%
  filter(GO.ID %in% reducedTerms_por_class$go)

#add in parent terms to list of go terms 
p05_clas_res_reduced_por$ParentTerm <- reducedTerms_por_class$parentTerm[match(p05_clas_res_reduced_por$GO.ID, reducedTerms_por_class$go)]

length(unique(p05_clas_res_reduced_por$ParentTerm))#14

p05_clas_res_reduced_por <- p05_clas_res_reduced_por %>% group_by(ParentTerm) %>% mutate("N_in_Parent" = dplyr::n()) %>% ungroup()
length(unique(p05_clas_res_reduced_por$ParentTerm)) #14
```

```{r}
write.csv(p05_clas_res_reduced_por, "output/por_parent_topGO_results_classicfisher.csv")
```

```{r}
parent_enrich_plot1_por_class <- ggplot(p05_clas_res_reduced_por, aes(x=classicFisher,y=reorder(Term,classicFisher)))+
  facet_grid(ParentTerm ~ ., scales="free", labeller = label_wrap_gen(width = 10, multi_line = TRUE))+
  geom_point(size=2, color="black")+
  geom_segment(aes(x=0, xend=classicFisher, y=Term, yend=Term)) +
  #geom_hline(yintercept = 0, linetype="solid", color = 'black', linewidth=0.5, show.legend = TRUE)+
  #scale_y_continuous(limits=c(0,40))+
   #scale_x_discrete(labels = label_wrap(30)) +
  #scale_fill_gradientn(colours=c("white","#fddbc7","#f4a582","#d6604d","#b2182b"), na.value = "grey98",limits = c(0, 100)) +
  #labs(title = "Enriched Biological Process Parent GO Terms, Brown Module", x="Parent Term", y="Number of Significant (p < 0.01) GO Terms in Module", colour="p-value") +
  theme_classic() + 
  theme(axis.text.y = element_text(size = 8), 
              axis.title = element_text(size = 8), 
              plot.title = element_text(hjust = 0.5, size = 8));parent_enrich_plot1_por_class

ggsave("figures/porites/parent_enrich_plot1_por_class.pdf", plot = parent_enrich_plot1_por, width = 8, height = 15, units = "in")
dev.off()  # close and save

```

```{r}
p05_clas_res_reduced_por$GeneRatio <- p05_clas_res_reduced_por$Significant / p05_clas_res_reduced_por$Annotated
parent_enrich_plot3_por_classic <- ggplot(p05_clas_res_reduced_por,
       aes(x = GeneRatio,
           y = reorder(Term, GeneRatio),
           color = -log10(classicFisher))) +
  geom_point(size=1) +
  geom_segment(aes(x=0, xend=GeneRatio, y=Term, yend=Term), color="grey70") +
  scale_color_gradient(low="lightblue", high="darkblue", name="-log10(classicFisher)") +
  theme_classic(base_size=10) +
  labs(x="Gene Ratio", y="GO Term") +
  theme(axis.text.y = element_text(size=5),
        axis.title = element_text(size=9),
        plot.title = element_text(hjust=0.5, size=10));parent_enrich_plot3_por_classic
ggsave(parent_enrich_plot3_por_classic, filename="figures/porites/por_parent_enrichment_individual_terms_classic.png", width=10, height=11)

#summarize by parent term
parent_summary_por_class <- p05_clas_res_reduced_por %>%
  group_by(ParentTerm) %>%
  summarise(
    maxLogP = max(-log10(classicFisher)),
    meanGeneRatio = mean(Significant/Annotated),
    nTerms = length(maxLogP)
  ) %>%
  arrange(desc(maxLogP))

parent_enrich_plot4_por_class<-ggplot(parent_summary_por_class, aes(x = meanGeneRatio, y = reorder(ParentTerm, meanGeneRatio), color = maxLogP, size = nTerms)) +
  geom_point(size=2) +
  scale_color_gradient(low = "lightblue", high = "darkblue", name="Max -log10(classicFisher)") +
  scale_size_continuous(name="Number of Terms") +
  theme_classic(base_size=10) +
  labs(x="Mean Gene Ratio", y="Parent GO Term");parent_enrich_plot4_por_class
ggsave(parent_enrich_plot4_por_class, filename="figures/porites/por_parent_enrichment_parent_terms_classic.png", width=10, height=7)
```